---
title: "Lab 4"
author: "Ram Balasubramanian, John Kenney"
date: "November 28, 2017"
output: pdf_document
---

```{r}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

library(xts)
```

#Description of the Lab

In this lab, you are provided a time series stored in "Lab4-series2.csv". The series, which is a monthly series,
covers the historical period from 1990 to 2015. Your main task is to build a time series model, using materials
covered between lecture 6 - lecture 10 (not including VAR modeling), and conduct a monthly, 11-month
ahead forecast of the series in 2015.

You will have to split the series into a training series, which includes the data from 1990 to 2014 December,
leaving all the months in 2015 as the test data.

As we have studied extensively in the last 6 lectures on how to build time series models, I expect that
your team illustrate and describe all of steps in building up to your final model. All of your analyses and
decisions to select your final models must be clearly demonstrated. It is quite possible that you will have
a few candidate models that you have considered. Besides using information criterion, consider the mean
absolute percentage error (MAPE) on the test sample in your model selection.

```{r}
#Load the data
df = as.data.frame(read.csv(
  'C:/Users/jkenney/Dropbox/UCB/Fall 2017/W271/Labs/Lab 4/Lab4_2017Fall/W271-Lab-4/Lab4-series2.csv', header = TRUE))
head(df$x)

#Create the Time Series
ts = as.xts(ts(df$x, frequency = 12, start = c(1990,1)))
str(ts)

#Separate into Test and Train Sets
train = ts['1990-01/2014-12']
tail(train)
test = ts['2015-01/']
test

#Autocorrelation
plot(train)
acf(train)
pacf(train)


View(pacf(ts[,2]))
install.packages("tseries")
library(tseries)

#Also look at stl()
decompose(ts[,2], type = c("additive"))

#Fourier Decomposition

adf.test(diff(ts[,2],1,1))

pacf(diff(ts[,2],1,1))
acf(diff(ts[,2],1,1))

dif.1 = diff(ts[,2],1,1)

pacf(diff(dif.1, lag = 12, differences = 1))
```



```{r echo=FALSE}

```
