---
title: "Lab 4"
author: "Ram Balasubramanian, John Kenney"
date: "November 28, 2017"
output: pdf_document
---

```{r}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

library(xts)
library(tseries)
library(lubridate)
```

#Description of the Lab

In this lab, you are provided a time series stored in "Lab4-series2.csv". The series, which is a monthly series,
covers the historical period from 1990 to 2015. Your main task is to build a time series model, using materials
covered between lecture 6 - lecture 10 (not including VAR modeling), and conduct a monthly, 11-month
ahead forecast of the series in 2015.

You will have to split the series into a training series, which includes the data from 1990 to 2014 December,
leaving all the months in 2015 as the test data.

As we have studied extensively in the last 6 lectures on how to build time series models, I expect that
your team illustrate and describe all of steps in building up to your final model. All of your analyses and
decisions to select your final models must be clearly demonstrated. It is quite possible that you will have
a few candidate models that you have considered. Besides using information criterion, consider the mean
absolute percentage error (MAPE) on the test sample in your model selection.

```{r}
x= as.data.frame(read.csv("Lab4-series2.csv", header = TRUE))
colnames(x) <- c("num", "value")
idx <- seq(as.Date("1990/1/1"), by = "month", length.out = dim(x)[1])

ts_all <- xts(x['value'], order.by=idx)
str(ts_all)
head(ts_all,10)
options(digits = 4)
#Create train set from 1990-2014
ts = ts_all['1990-01-01/2014-12-31']

  
```
# Exploratory Data Analysis:

## Plot the time series
```{r}
plot.xts(ts, type='l', minor.ticks = NULL, major.ticks = 12, main="Time Series Plot", grid.ticks.on="years")
```

## Observations:  
Trend:  Series doesn't show any overall trend;  there appear to be periods of "shock" where the series goes up quickly and slowly dissipates from those highs. In general the series exhibits "random walk" behavior.  
Seasonal & Cyclical patterns:  There are some seasonal patterns apparent in the series. There is also a broad cyclical pattern (ups and downs) - but the periodicity doesn't appear to be consistent.   
Outliers:  There aren't any apparent outliers.


## Check for Stationarity
```{r}
#is the data stationary in the mean/variance?
#Let's look at the acf and pacf plots
  par(mfrow=c(1,2))
  acf(ts, main="ACF of Given Series")
  pacf(ts, main="PACF of Given Series")
```
## Observations:  
The ACF declines very slowly (indicative of an AR(1) process).  Infact the ACF signature looks like a random walk process.
The PACF shows significance at lag-1.  There is also significant correlation at lags 4,7,10,11,13. Note that corellation at lag 13 is negative. 



```{r echo=FALSE}

#Conduct Jlung-Box test to see if data are serially correlated
  BoxPValue = Box.test(ts,type='Ljung-Box',lag=20)$p.value
  result = "Data are Random"
  if (BoxPValue < 0.05){ result = "Data are NOT random "} 
  print(paste(result, " - p-value of box-test = ",BoxPValue))

#conduct Dicky-Fuller test to see if series is stationary
  adfPvalue = adf.test(ts)$p.value
  result = "Unit Root Present - Non Stationary Series"
  if (adfPvalue < 0.05){ result = " Reject Unit Root Presence - Stationary Series"} 
  print(paste(result, " - p-value of adf-test =",adfPvalue))

#monthplot(ts)

```
## Dicky-Fuller test shows that the series is not stationary.  While there is no clear trend in the data - it does exhibit random-walk behavior.  Let's apply a first difference to see if that makes the series stationary.



## First Difference
```{r}
diff1 = diff(ts, lag=1, differences=1)['1990-02-01/'] #get rid of the first NA
autoplot.zoo(diff1)

par(mfrow=c(1,2))
acf(diff1, na.action=na.pass)
pacf(diff1, na.action=na.pass)

adfPvalue = adf.test(diff1)$p.value

result = "Unit Root Present - Non Stationary Series"
if (adfPvalue < 0.05){ result = " Reject Unit Root Presence - Stationary Series"} 
print(paste(result, " - p-value of adf-test =",adfPvalue))

```
Observations:
Taking the first difference makes the series stationary.  But the plot of the series and the ACF shows a strong seasonal pattern. The ACF plot shows strong  correlations at lag 3 (negative), 6 (postive), 9 (-ve), 12 (+ve).  The PACF shows similar seasonal patterns. 
Let's investigate that further. 


## Investigate seasonality
```{r}
#Look at monthly averages
par(mfrow=c(1,2))
plot(aggregate(ts, month(index(ts)), mean), type='o')
monthplot(ts)
```
### The month plot shows there are "high seasons" and "low seasons".
```{r}
#since this is a monthly series, let's try a 12-m seasonal difference

diff12 = diff(ts, lag=12, differences = 1)['1991-01-01/']
autoplot.zoo(diff12)
par(mfrow=c(1,2))
acf(diff12)
pacf(diff12)
```

## The acf shows a slow decline, and the pacf plot shows a strong drop after lag-1 - both characteristics of an AR-1 process.  The pacf also shows strong correlation at Lags 3, 6, and at 13 (WHAT IS THIS SUPPOSED TO MEAN???)


```{r}
adfPvalue = adf.test(diff12)$p.value
result = "Unit Root Present - Non Stationary Series"
if (adfPvalue < 0.05){ result = " Reject Unit Root Presence - Stationary Series"} 
print(paste(result, " - p-value of adf-test =",adfPvalue))

```

#At this point we are looking at a ARIMA(1,0,0),(0,1,0)_12 model.

```{r}

#JUST PLAYING WITH OTHER LAGS IN ADDITION TO 12.  SOMEWHAT LOST HERE

diff3 = diff(diff12, lag=3, differences=1)['1991-04-01/']
acf(diff3)
pacf(diff3)
```

